---
title: Course Project 2
author: "Hesham Elhalawani"
date: "June 25, 2017"
output:
  pdf_document: default
  html_document: default
---
# Health and Economic Impact of Weather Events in the US


Storms and other severe weather events can cause both public health and economic
problems for communities and municipalities. Many severe events can result in
fatalities, injuries, and property damage, and preventing such outcomes to the extent
possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric
Administration's (NOAA) storm database. This database tracks characteristics of major
storms and weather events in the United States, including when and where they occur, as
well as estimates of any fatalities, injuries, and property damage.

Synopsis
========

The analysis on the storm event database revealed that tornadoes are the most
dangerous weather event to the population health. Whereas flash floods were responsible for the worst property damages between 1950 and 2011.


Data Processing

The analysis was performed on
[Storm Events Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp), provided by
[National Climatic Data Center](http://www.ncdc.noaa.gov/). The data is from a comma-separated-value file available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).
There is also some documentation of the data available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).


The first step is to read the data into a data frame.
```{r cache=TRUE}
storm <- read.csv(bzfile("repdata%2Fdata%2FStormData.csv.bz2"))
```

Before the analysis, the data need some preprocessing, especially regarding standardizing the format of event types, i.e. upper/lower cases.


```{r}
# number of unique event types
length(unique(storm$EVTYPE))
# translate all letters to lowercase
event_types <- tolower(storm$EVTYPE)
# replace all punct. characters with a space
event_types <- gsub("[[:blank:][:punct:]+]", " ", event_types)
length(unique(event_types))
# update the data frame
storm$EVTYPE <- event_types
```

For further analysis, the cleaned event types are used.


## Dangerous Events with respect to Population Health


To find the event types that are most harmful to population health, the number
of casualties are aggregated by the event type.

```{r}
library(plyr)
casualties <- ddply(storm, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))

# Find events that caused most death and injury
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
injury_events <- head(casualties[order(casualties$injuries, decreasing = T), ], 10)
```

Top 10 events that caused largest number of deaths are

```{r}
fatal_events[, c("EVTYPE", "fatalities")]
```

Top 10 events that caused most number of injuries are

```{r}
injury_events[, c("EVTYPE", "injuries")]
```

## Economic Effects of Weather Events


To analyze the impact of weather events on the economy, available property
damage and crop damage reportings/estimates were used. The first step in the
analysis is to calculate the property and crop damage for each event.

```{r}
exp_transform <- function(e) {
    # h -> hundred, k -> thousand, m -> million, b -> billion
    if (e %in% c('h', 'H'))
        return(2)
    else if (e %in% c('k', 'K'))
        return(3)
    else if (e %in% c('m', 'M'))
        return(6)
    else if (e %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(e))) # if a digit
        return(as.numeric(e))
    else if (e %in% c('', '-', '?', '+'))
        return(0)
    else {
        stop("Invalid exponent value.")
    }
}
```

```{r cache=TRUE}
prop_dmg_exp <- sapply(storm$PROPDMGEXP, FUN=exp_transform)
storm$prop_dmg <- storm$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- sapply(storm$CROPDMGEXP, FUN=exp_transform)
storm$crop_dmg <- storm$CROPDMG * (10 ** crop_dmg_exp)
```


```{r}
# Compute the economic loss by event type
library(plyr)
econ_loss <- ddply(storm, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))

# filter out events that caused no economic loss
econ_loss <- econ_loss[(econ_loss$prop_dmg > 0 | econ_loss$crop_dmg > 0), ]
prop_dmg_events <- head(econ_loss[order(econ_loss$prop_dmg, decreasing = T), ], 10)
crop_dmg_events <- head(econ_loss[order(econ_loss$crop_dmg, decreasing = T), ], 10)
```

Top 10 events that caused most property damage (in dollars) are as follows

```{r}
prop_dmg_events[, c("EVTYPE", "prop_dmg")]
```

Similarly, the events that caused biggest crop damage are

```{r}
crop_dmg_events[, c("EVTYPE", "crop_dmg")]
```

## Results

### Health impact of weather events


The following plot shows top dangerous weather event types.

```{r}
library(ggplot2)
library(gridExtra)
# Set the levels in order
p1 <- ggplot(data=fatal_events,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities), fill=fatalities) 
    geom_bar(stat="identity") +
    coord_flip() +
    ylab("Total number of fatalities") +
    xlab("Event type") +
    theme(legend.position="none")

p2 <- ggplot(data=injury_events,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    ylab("Total number of injuries") +
    xlab("Event type") +
    theme(legend.position="none")

grid.arrange(p1, p2)
```

Tornadoes cause most number of deaths and injuries among all event types. 


## Economic impact of weather events
---------------------------------

The following plot shows the most severe weather event types with respect to
economic cost that they have costed since 1950s.


```{r}
library(ggplot2)
library(gridExtra)
# Set the levels in order
p1 <- ggplot(data=prop_dmg_events,
             aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
    geom_bar(stat="identity") +
    coord_flip() +
    xlab("Event type") +
    ylab("Property damage in dollars (log-scale)") +
    theme(legend.position="none")

p2 <- ggplot(data=crop_dmg_events,
             aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    xlab("Event type") +
    ylab("Crop damage in dollars") + 
    theme(legend.position="none")

grid.arrange(p1, p2)
```


The most severe weather event in terms of crop damage is the drought, followed by floods and hails.